When does vapor pressure deficit drive or reduce evapotranspiration?

### 2019-02-27 doc/paper/body.tex
need to find good citiations for ESM/LSM claim, email dan?

By varying four parameters (g$_a$, T, uWUE, g$_1$) at three different
values (Table \ref{param_varying}) we generate nine different values
for the scaling term, nine different curves (as a function of VPD) for
the sign term, and 81 different curves for the ET response to VPD
($\frac{\partial \; ET}{\partial \; VPD}$). Examining a subset of nine
of the curves for the ET response to VPD, defined by the minimum, median
and maximum of for each of the scaling and sign term, it is clear
that the range of responses includes those where ET almost always
decreases with increasing VPD (water conservative), ET almost always
increases with increasing VPD (water intensive), and intermediate
values where ET will increase or decrease with VPD depending on the
atmospheric demand for water (Figure \ref{full}).

@@

Examining the sign term independent of the scaling term illuminates
the role of the two plant parameters, g$_1$ and uWUE, in determining
the degree to which the response is water conservative
($\frac{\partial \; ET}{\partial \; VPD} < 0$) or water intensive
($\frac{\partial \; ET}{\partial \; VPD} > 0$) (Figure \ref{sign})
. Higher g$_1$ and uWUE shift the curve towards increasing ET
responses with VPD (water intensive), and smaller g$_1$ values lead to
a higher non-linear VPD dependence of the response.

@@

@@

Interpreting both the derivation and results of this manuscript hinges
on an understanding and appreciation for the usefulness of partial
derivatives for understanding the behavior of complex systems. Our
results are a conceptual tool for answering the question, ``What is
the ET response to VPD with all other quantities held fixed?'' This is
useful, and we would argue is critical, for understanding complex
responses when many quantities are varying simultaneously. If we
cannot understand the contribution of one term to observed
variability, independent of other confounding factors, there is little
hope of disentangling and understanding the relative role of many
quantities in determining observed variability.  Equation \ref{d_et}
explicitly provides an estimate of the ET response to VPD, given
assumptions about ecosystem plant characteristics through the
parameters g$_1$ and uWUE. Eq. (\ref{d_et}) is generic to external
environmental factors and the timescale over which a VPD response
acts, and has a straightforward conceptual meaning. It provides the ET
response, given other quantities and parameters held fixed, subject to
the assumptions outlined in Sect. \ref{methods}. So, while the plant
parameters g$_1$ and uWUE may vary with plant species and
environmental conditions like soil moisture, we can still assess the
ET response to VPD for a given plant species, or for a given soil
moisture condition.

However, given that g$_1$ and uWUE can vary, and they both have a
large impact on the ET response (Figure \ref{full}, \ref{sign}), it is
useful to examine how these quantities vary in previous literature,
framed by our results on $\frac{\partial \; ET}{\partial \; VPD}$. In
\citet{Zhou_2015} most plant types have uWUE similar to our median
value, under which the sign of the ET response depends on
VPD. However, crops, which we expect to prioritize carbon uptake over
water conservation, have a higher uWUE closer to our maximum
value. This higher uWUE results in a higher likelihood for an
increasing ET response to VPD, which intuitively makes sense given
that we expect crops to keep stomata open for access to carbon, at the
cost of increased water loss during high VPD. Shrubs, and to a lesser
extent grass, have an uWUE closer to the minimum uWUE. For these plant
types, we would then expect a higher likelihood of a water
conservative response; that is, a decrease of ET in response to
VPD. It is important to note that, while \citet{Zhou_2015} did not
examine the role of soil moisture for within plant-type variability of
uWUE, we might expect some variation, especially in extreme cases when
soil water becomes a limiting factor.

In \citet{Medlyn_2017} there is not as robust of a relationship
between plant type and g$_1$ as there was for uWUE in
\citet{Zhou_2015}. Given that there is a robust relationship between
g$_1$ and plant type in \citet{Lin_2015}, the ambiguous results in
\citet{Medlyn_2017} could be due to noise in ecosystem scale
observations, the consequences of imposing a model structure in some
of their g$_1$ esimations, or real ecosystem-scale within-plant type
variability. These differences between leaf-scale behavior and
ecosystem scale behavior highlight the importance of understanding how
and why slope parameters show increased variability at the ecosystem
scale, as argued by \citet{Medlyn_2017}. However, if we assume that
there is some real analogy between leaf scale results in
\citet{Lin_2015} and ecosystem scale behavior, and that some of the
ambiguity in \citet{Medlyn_2017} is due to model and/or observational
error, then we can make some connections between g$_1$-induced ET
response variability and plant characteristics, which are subject to
the strong caveat that we still do not fully understand g$_1$'s
utility at the ecosystem scale. Extrapolating \citet{Lin_2015}'s
results (e.g. Fig. 2) to the ecosystem scale would reveal the
following relationships between g$_1$ and plant types:

\begin{itemize}
  \item C3 plants would have a generally higher g$_1$ than C4 plants,
  \item Crops would have a larger g$_1$ than shrubs, grass, and
    angiosperm trees, which have a higher g$_1$ than gymnosperm
    trees.
  \item Tropical and temperate climates would generally host plants
    with a higher g$_1$ than arctic and boreal climates.
\end{itemize}
For all of the above relationships, a higher g$_1$ means a higher
likelihood of a positive ET response to increasing VPD (water
intensive), and a smaller VPD dependence of the ET response.

Our theoretical results, when framed by previous literature tying
parameters to plant types and climate, highlight the variability of ET
response to VPD as a function of plant characteristics. Whether an
ecosystem increases or decreases ET in response to VPD will vary
depending on vegetation and climate. Generalizations about ET response
to VPD therefore require thoughtful consideration of ecosystem
characteristics; statements such as ``ET increases with warming due to
increases in VPD'' may be false depending on the water conservation
strategy of a given ecosystem **CITE SOME EXAMPLES OF THIS LOGIC FROM
GLOBAL WARMING LIT**.

Additionally, the sensitivity of ET response to plant parameters
highlights the importance of understanding and developing stomatal
conductance models and parameters for the ecosystem scale. In
particular, why and how g$_1$ behavior at the leaf and ecosystem scale
are not directly analogous much be understood. While the derivation
here was explicit about the assumption of a constant g$_1$ term with
respect to the VPD perturbation, many sophisticated land surface
models and earth system models employing stomatal conductance models
make similar, and even stronger assumptions about the stationarity of
VPD slope terms. Many models will assume a constant g$_1$ term within
a given plant functional type, and research derived from these models
often does not acknowledge limitations of this assumption, or the
difficulty in the literature with quantifying an ecosystem scale slope
parameter \citep{Medlyn_2017}. Given that something so fundamental as
the ET response to VPD varies strongly with g$_1$, we must invest more
in understanding g$_1$'s behavior at the ecosystem scale, and the
efficacy of models using g$_1$'s for ecosystem-scale
fluxes. Fortunately, our framework, and the novel general approach of
using robust ecosystem-scale ratios (uWUE) to remove the implicit
dependence of stomatal conductance on ET, is flexible enough to be
applied to any future-derived ecosystem-scale stomatal conductance
model that contains a dependence on GPP.

>>>>

By varying four parameters (g$_a$, T, uWUE, g$_1$) at three different
values (Table \ref{param_varying}) we generate nine different values
for the scaling term, nine different curves (as a function of VPD) for
the sign term, and 81 different curves for the ET response to VPD
($\frac{\partial \; ET}{\partial \; VPD}$). To aid visualization we
can examine a subset of nine of these 81 curves, defined by the
minimum, median and maximum values for both the scaling and sign term
(Figure \ref{full}). The range of ET responses to VPD vary from those
where ET almost always decreases with increasing VPD (water
conservative), to those where ET almost always increases with
increasing VPD (water intensive). Additionally, for some parameters
whether ET will increase or decrease with increasing VPD depends on
atmospheric demand (Figure \ref{full}).

@@

@@

Examining the sign term independent of the scaling term illuminates
the role of the two plant parameters, g$_1$ and uWUE, in determining
the degree to which the response is water conservative
($\frac{\partial \; ET}{\partial \; VPD} < 0$) or water intensive
($\frac{\partial \; ET}{\partial \; VPD} > 0$) (Figure \ref{sign})
. Higher g$_1$ and uWUE shift the curve towards increasing ET
responses with VPD (water intensive), and smaller g$_1$ values lead to
a higher non-linear VPD dependence of the response. g$_1$ and uWUE
should generally co-vary with $\lambda$. As $\lambda$ increases, g$_1$
should increase and uWUE should decrease. ** ellaborate **

@@

Interpreting both the derivation and results of this manuscript hinges
on an understanding and appreciation for the usefulness of partial
derivatives for understanding the behavior of complex systems in a
simplifed framework. Our results are a conceptual tool for answering
the question, ``What is the ET response to VPD with all other
quantities held fixed?'' This is useful, and we would argue is
critical, for understanding complex responses when many quantities are
varying simultaneously \citep{Zhou_2019}. If we cannot understand the
contribution of one term to observed variability, independent of other
confounding factors, there is little hope of disentangling and
understanding the relative role of many covarying quantities in
determining observed variability.  Equation \ref{d_et} explicitly
provides an estimate of the ET response to VPD, given assumptions
about ecosystem plant characteristics through the parameters g$_1$ and
uWUE. Eq. (\ref{d_et}) is generic to external environmental factors
and the timescale over which a VPD response acts, and has a
straightforward conceptual interpretation. It provides the ET
response, given other quantities and parameters held fixed, subject to
the assumptions outlined in Sect. \ref{methods}. So, while the plant
parameters g$_1$ and uWUE may vary with plant species and
environmental conditions like soil moisture, we can still assess the
ET response to VPD for a given plant species, or for a given soil
moisture condition. ** talk about uWUE-SM, with supplementary here **?

However, given that g$_1$ and uWUE can vary, and they both have a
large impact on the ET response (Figure \ref{full}, \ref{sign}), it is
useful to examine how these quantities vary in previous literature,
framed by our results on $\frac{\partial \; ET}{\partial \; VPD}$. In
\citet{Zhou_2015} most plant types have uWUE similar to our median
value, under which the sign of the ET response depends on
VPD. However, crops, which we expect to prioritize carbon uptake over
water conservation, have a higher uWUE value closer to our maximum
value. This higher uWUE results in a higher likelihood for an
increasing ET response to VPD, which matches intuition given that we
expect crops to keep stomata open for access to carbon, at the cost of
increased water loss during high VPD. Shrubs, and to a lesser extent
grass, have an uWUE closer to the minimum uWUE. For these plant types,
we would then expect a decrease of ET in response to VPD. It is
important to note that, while \citet{Zhou_2015} did not examine the
role of soil moisture for within plant-type variability of uWUE, we
might expect some variation, especially in extreme cases when soil
water becomes a limiting factor. ** talk about uWUE-SM, with supplementary here **?

\citet{Medlyn_2017}'s results did not display the robust relationship
between plant type and and plant parameter (g$_1$) that was
demontrated by \citet{Zhou_2015} for uWUE. Given that there is a
robust relationship between g$_1$ and plant type in \citet{Lin_2015},
and that \citet{Zhou_2015}'s ecosystem-scale results rely on the same
underlying theory as \citet{Medlyn_2017}'s results, the ambiguous
results in \citet{Medlyn_2017} could be due to noise in ecosystem
scale observations, the consequences of imposing a model structure in
some of their g$_1$ esimations, or real ecosystem-scale within-plant
type variability (e.g. multiple species impacting the relationship
with the eddy-covariance footprint). These differences between
leaf-scale behavior and ecosystem scale behavior highlight the
importance of understanding how and why slope parameters show
increased variability at the ecosystem scale, as argued by
\citet{Medlyn_2017}. However, if we assume that there is some real
analogy between leaf scale results in \citet{Lin_2015} and ecosystem
scale behavior, and that some of the ambiguity in \citet{Medlyn_2017}
is due to model and/or observational error, then we can g$_1$-induced
ET response variability and plant characteristics, which are subject
to the strong caveat that we still do not fully understand g$_1$
behavior at the ecosystem scale. Extrapolating \citet{Lin_2015}'s
results (e.g. Fig. 2) to the ecosystem scale would reveal the
following relationships between g$_1$ and plant types:

\begin{itemize}
  \item C3 plants would have a generally higher g$_1$ than C4 plants,
  \item Crops would have a larger g$_1$ than shrubs, grass, and
    angiosperm trees, which have a higher g$_1$ than gymnosperm
    trees.
  \item Tropical and temperate climates would generally be
    characterized by plants with a higher g$_1$ than arctic and boreal
    climates.
\end{itemize}
For all of the above relationships, a higher g$_1$ means a higher
likelihood of a positive ET response to increasing VPD (water
intensive), and a smaller VPD dependence of the ET response.

Our theoretical results highlight the variability of ET
response to VPD as a function of both climate  and plant
characteristics, especially water usage strategy. Whether an
ecosystem increases or decreases ET in response to VPD will vary
depending on vegetation and climate. Generalizations about ET response
to VPD therefore require thoughtful consideration of ecosystem
characteristics; statements such as ``ET increases with warming due to
increases in VPD'' may be false depending on the water conservation
strategy of a given ecosystem \citep{Lemordant_2018}.

Additionally, the sensitivity of ET response to plant parameters
highlights the importance of understanding and developing stomatal
conductance models and parameters for the ecosystem scale. In
particular, why and how g$_1$ behavior at the leaf scale is not
analagous to the ecosystem scale must be understood. While the derivation
here was explicit about the assumption of a constant g$_1$ term with
respect to the VPD perturbation, many sophisticated land surface
models and earth system models employing stomatal conductance models
make similar, and even stronger assumptions about the stationarity of
VPD slope terms. Many models will assume a constant g$_1$ term within
a given plant functional type, and research derived from these models
often does not acknowledge limitations of this assumption, or the
difficulty in the literature with quantifying an ecosystem scale slope
parameter \citep{Medlyn_2017}. Given that something so fundamental as
the ET response to VPD varies strongly with g$_1$, we must invest more
in understanding g$_1$'s behavior at the ecosystem scale, and the
efficacy of models using g$_1$'s for ecosystem-scale
fluxes. Fortunately, our framework, and the novel approach of
using robust ecosystem-scale ratios (uWUE) to remove the implicit
dependence of stomatal conductance on ET, is flexible enough to be
applied to any future-derived ecosystem-scale stomatal conductance
model that contains a dependence on GPP.
